Normalization: A pipeline to analyse time-course gene expression data

### 2019-02-16 scripts/01_quality_control_normalization.Rmd
Now normalizing data as part of the paper

>>>>

```{r echo=FALSE, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  cache=FALSE, autodep=FALSE, warning=FALSE, error=FALSE, message=FALSE,
  echo=FALSE,
  duplicate.label="allow"
)

library(genefilter)
library(dplyr)
library(ggplot2)
library(NMF)
library(reshape2)
library(limma)
```

The paper describes a normalization strategy that I am going to use here:

  - First, normalize probe intensity using background information using the
    "norm-exponential" strategy.
  - Then normalize across micro-arrrays using quantily normalization.

Unfortunately, that means going back to the raw data, which can be downloaded
in the data folder, by clicking on the link specified in the README (it might
take a while…). Untar the main file into a `rawdata` folder.

```{r reading_targets, echo=FALSE, include=FALSE}

targets = readTargets()
RG = read.maimages(
  targets, source="agilent",
  path="data/shoemaker2015/rawdata",
  other.columns="gIsWellAboveBG",
  green.only=TRUE)

```

Before quantile normalization:

```{r echo=FALSE, include=FALSE}
RG = backgroundCorrect(RG, method="normexp")
```

```{r plot_densities, echo=FALSE}
xlim = c(0, 14)
plotDensities(RG, col="black", legend=FALSE)
```

After quantile normalization:

```{r normalize_between_arrays, echo=FALSE}
RG.q = normalizeBetweenArrays(RG, method="quantile")
plotDensities(RG.q, col="black", legend=FALSE)

sample_names = sapply(strsplit(colnames(RG.q), "_"), .subset2,  1)
colnames(RG.q) = sample_names
row.names(RG.q) = make.names(RG$genes$SystematicName, unique=TRUE)
idx_to_keep = which(
  !(RG$genes$Systematic %in% c("DarkCorner", "GE_BrightCorner",
			       "NegativeControl")))

data = RG.q[idx_to_keep,]

# Filter out control probes.
control_probes = data$genes[, "ControlType"] == 1
is_expressed = rowSums(data$other$"gIsWellAboveBG" > 0) >= 4

data = data[!control_probes & is_expressed, ]$E

write.table(data, "results/quantile_normalized.txt")

meta = read.delim("data/shoemaker2015/mice_meta.tsv", sep="\t")

```

```{r order_genes, echo=FALSE}
# Reorder genes on condition, time, and replicate
ord = order(
  meta$Group,
  meta$Time,
  meta$Replicate)
data = data[, ord]
meta = meta[ord, ]
meta["Time"] = as.factor(unlist(meta["Time"]))
meta["Replicate"] = as.factor(unlist(meta["Replicate"]))
```

## Plot the RLE and PCA

```{r leafRleRaw, fig.width=6}
library(EDASeq)
ylim = c(-2.5,2.5)
plotRLE(exp(data), outline=FALSE,
        whisklty=1, whisklwd=2, col=meta$Replicate, main="RLE, data")
plotRLE(exp(data), outline=FALSE,
        whisklty=1, whisklwd=2, col=meta$Group, main="RLE, data")
plotRLE(exp(data), outline=FALSE,
        whisklty=1, whisklwd=2, col=meta$Time, main="RLE, data")

```
```{r leafPCARaw}
library(ggfortify)
pca_data = prcomp(t(data), rank=5, center=TRUE, scale=TRUE) 
percent_var = round(100 * attr(pca_data, "percentVar"))

autoplot(pca_data, data=meta, colour="Group")
autoplot(pca_data, data=meta, colour="Time")
autoplot(pca_data, data=meta, colour="Replicate")

```

```{r fig.width=6, fig.height=6}
library(NMF)

data_corr = cor(data, method="pearson")
aheatmap(data_corr,
         Colv=NA, Rowv=NA,
         annCol=meta[,
		     c("Group", "Time",
			"Replicate")],
         annLegend=FALSE, 
         main="Replicates\nAfter Normalization")
```

```{r fig.width=6, fig.height=6}
library(NMF)

data_corr = cor(data, method="spearman")
aheatmap(data_corr,
         Colv=NA, Rowv=NA,
         annCol=meta[,
		     c("Group", "Time",
			"Replicate")],
         annLegend=FALSE, 
         main="Replicates\nAfter Normalization")
```
